Bayesian exponential random graph modeling of KWL Burial network

### 2020-11-13 analysis/paper/paper.Rmd
editing the text together

<!-- Bayesian inference for Exponential Random Graph Models --> 
Bayesian approaches to ERGMs are efficient computational tools for social network analysis because they incorporates prior information about the network configurations into the model and offers uncertainty quantification by evaluating the posterior distributions of the parameters associated with network configurations [@Caimo2015; @Nemmers2019]. The advantage that Bayesian modeling has over traditional ERGMs is the application of Markov chain Monte Carlo (MCMC) simulation using the approximate exchange algorithm [@Caimo2011]. MCMC avoids doubly-intractable computations by directly sampling from the not normalized part of the posterior, which alleviates the computational problems and gives a better convergence results. This enable us to deal with complicated dependence patterns with ease, providing better estimations for complex social network models with heterogeneous data [@Caimo2017; @Snijders2006]. By fitting an ERMG with an MCMC algorithm, a Bayesian approach generates posterior probabilities that incorporates our sample data and prior information through summary statistics from estimates of the ERGM parameters. Posterior probability estimates the effect of ERGM parameters by looking at the posterior mean and 95% credible intervals, which can replace p-values for assessing a null hypothesis [@Caimo2017bayesian]. The typical criteria for interpreting the posteriors is that an odds ratio greater than one means a positive effect of a parameter, while odds ratios less than one represents negative effect. In addition, Bayesian approaches are also useful to deal with missing data, which is often a problem leading to misinterpretation of networks, especially for archaeological studies. @Koskinen2010 shows that the effect of missing data can be reduced with Bayesian modeling that can predict, on the average, 80% of the ties when a third of data missing.

@@

<!-- we should tell the reader what does it mean high-medium-low --> 
We analyzed burial data collected from the published excavation reports, and the original fieldwork notes of Kiwulan. Gold-foil and agate beads are commonly found across burials with substantial differences in quantities, so we described each burial as having one of three levels, high (> 10), medium (> 1 and < 10), and low (= 1), of these beads. Glass beads are in levels of high (> 100), medium (> 99 and < 900), and low (< 100). That means if burial 1 and burial 2 both have high quantities of agate beads, then there will be a tie connecting them. For less frequent prestige goods, including imported porcelains, gold foils, and fish-shaped ornaments, we linked two burials when they possess each type of goods in common (i.e. presence or absence). Node attributes here include osteological data, such as age and sex, and cultural data, such as ritual pottery, and burial value index. Ritual pottery was identified as locally made ceramics placed above graves that suggests funeral feasting according to historical records. The burial value index is an important attribute for economical inequality that we assigned the burials into three classes, high (>30), medium(>11 and <30), and low(<12), as an index of wealth. The number is the sum of values of each type of prestige good, which is calculate by the total number of a prestige item from burial context over the number of prestige item in a burial. Since burials tend to have multiple prestige goods in common, the network ties are weighted instead of binary data (where the value 1 represents a tie and the value 0 otherwise) [@Snijders2011]. For example, if two burials have both low quantity of glass beads and porcelain in common, the tie is given a value of 2. Our networks are non-directed, which means ties have no orientation forming the relationship between actors to indicate a mutual relationship.

@@

Our pericolonial impact hypothesis is that restricted grave goods or accumulation of goods present in a few individuals after the European presence in the 17th century, because of increased social inequality at Kiwulan due to differential access high value goods. The high value of trade goods had been identified by the historical documents written by the Europeans and the Han Chinese [@Borao2009; @LiandWu2006]. In addition to the interest of burial goods, the indirect colonialism in northeastern Taiwan may suggest the potentially manipulation of the European colonial image by local ambitious individuals for building their personal status and power. If the social inequality gradually increased in Kiwulan, then we would observe a network with higher degree (popularity), less transitivity (cohesion), and strong inequality based on age difference (achieved status). Network analysis provides a way to examine the underlying mechanisms that shape the relationships between people at Kiwulan, and the effect of foreign contact on these relationships through chronological comparison.

## Model specification in a Bayesian framework
```{r ergm-terms-table}
ergm_terms <- readxl::read_excel(here("analysis", "data", "raw_data", "ergm_term.xlsx")) 
kableExtra::kable(ergm_terms,
             caption = "The parameters of exponential random graph models for a undirected network to corresponding with archaeological evidence.")
```

<!--Network construction and specification-->
We used Bayesian inference on exponential random graph models (ERGMs) to quantify the relations among burials and test our hypothesis of social change using the R programming language with the bergm package [@Caimo2014], which built upon the statnet package for model specification and simulation procedures [@Hunter2008; @Morris2008; @Handcock2008]. Table \@ref(tab:ergm-terms-table) lists the configurations we used for the model specification for burials with archaeological interpretations. Every parameter in an ERGM has an associated algorithm for computing its value for a network [@Morris2008]. Based on our hypothesis, we model a network with increased social inequality that represented by low reciprocity, low transitivity, high popularity, high age homophily, and less proximity structure effects. We modeled several node attributes for homophily effects, including age, gender, ritual, and the degree of wealth, to explore the correlation between the tendency of connected nodes and the network ties overall. We also take into account the physical distance between burials as an indicator of a kinship-based relation since the deceased from the same family tend to be buried close to each other. Our model infers the presence of a few individuals as centers having more relations with others, which hints the emergence of social inequality. We expect that observed network after the European arrival will have a better match with our model compared with the network before European contact.

<!--Model fitting and prior setting--> 
After the specification, the network is simulated in a Baysesian framework using Markov chain Monte Carlo (MCMC) algorithm. MCMC algorithm is an estimation of posterior distribution of interest through directly random sampling the posterior without assuming the prior is a normal distribution. We can obtain a distribution by constructing a Markov chain that describes a sequence of moves from current state to the next state following probabilistic rules based on algorithms. This enables a random or stochastic simulation in a long run without depending on the previous move. More chains ensures a more desirable distribution close to the target distribution that we study, which is called the convergence of the Markov chains. In Bayesian ERGMs, MCMC first selects a set of edges (or a set of empty pair of actors) with equal probability, and than switch to a pair of actors at random within the chosen set [@Caimo2011]. In our case, we set the number of chains to six. For each chain, the number of burn-in iterations is 200 and the number of iterations after the burn-in is 2000. We set the number of iterations used to simulate a network y′ at each iteration to 10,000. For prior, the normal distribution is agreed as a suitable prior model in network analysis that assumes networks with low density and high transitivity commonly found in real world [@Caimo2017].

<!-- Goodness-of-fit assessment-->
The models are assessed by goodness-of-fit (GOF) diagnostics in the Bayesian framework, where the observed network is compared with the set of networks simulated from the estimated posterior distribution of the parameters of the model [@Caimo2011; @Caimo2017]. The Bayesian GOF summarized three distributions, including degree, minimum geodesic distance, and edgewise shared partner distributions. This provides a statistical approach to check how well the estimated posterior parameter distribution based on our hypotheses can reproduce networks similar to the general structural features of the observed network. We then compared the distribution of our observed networks, pre-European and post-European networks, with the distribution of our hypothesized model with higher popularity. We expect to see the network after the European presence has a better fit with the hypothesis, indicative of a structure with higher social differentiation.

>>>>

<!-- Bayesian inference for Exponential Random Graph Models --> 
Bayesian approaches to ERGMs are efficient computational tools for social network analysis because they incorporates prior information about the network configurations into the model and offers uncertainty quantification by evaluating the posterior distributions of the parameters associated with network configurations [@Caimo2015; @Nemmers2019]. <!-- add brief sentence to define posterior and prior to help the reader --> The advantage that Bayesian modeling has over traditional ERGMs is the application of Markov chain Monte Carlo (MCMC) simulation using the approximate exchange algorithm [@Caimo2011]. MCMC avoids doubly-intractable computations by directly sampling from the not normalized part of the posterior, which alleviates the computational problems and gives a better convergence results. This enable us to deal with complicated dependence patterns with ease, providing better estimations for complex social network models with heterogeneous data [@Caimo2017; @Snijders2006]. By fitting an ERMG with an MCMC algorithm, a Bayesian approach generates posterior probabilities that incorporates our sample data and prior information through summary statistics from estimates of the ERGM parameters. Posterior probability estimates the effect of ERGM parameters by looking at the posterior mean and 95% credible intervals, which can replace p-values for assessing a null hypothesis [@Caimo2017bayesian]. The typical criteria for interpreting the posteriors is that an odds ratio greater than one means a positive effect of a parameter, while odds ratios less than one represents negative effect. In addition, Bayesian approaches are also useful to deal with missing data, which is often a problem leading to misinterpretation of networks, especially for archaeological studies. @Koskinen2010 shows that the effect of missing data can be reduced with Bayesian modeling that can predict, on the average, 80% of the ties when a third of data missing.

@@

<!-- we should tell the reader what does it mean high-medium-low --> 
We analyzed burial data collected from the published excavation reports, and the original fieldwork notes of Kiwulan. Gold-foil and agate beads are commonly found across burials with substantial differences in quantities, so we described each burial as having one of three levels, high (> 10), medium (> 1 and < 10), and low (= 1), of these beads. Glass beads are in levels of high (> 100), medium (> 99 and < 900), and low (< 100). That means if burial 1 and burial 2 both have high quantities of agate beads, then there will be a tie connecting them. For less frequent prestige goods, including imported porcelains, gold foils, and fish-shaped ornaments, we linked two burials when they possess each type of goods in common (i.e. presence or absence). Node attributes here include osteological data, such as age and sex, and cultural data, such as ritual pottery, and burial value index. Ritual pottery was identified as locally made ceramics placed above graves that suggests funeral feasting according to historical records. The burial value index is an important attribute for economical inequality that we assigned the burials into three classes, high (>30), medium (>11 and <30), and low (<12), as an index of wealth. The number is the sum of values of each type of prestige good, which is calculate by the total number of a prestige item from burial context over the number of prestige item in a burial. Since burials tend to have multiple prestige goods in common, the network ties are weighted instead of binary data (where the value 1 represents a tie and the value 0 otherwise) [@Snijders2011]. For example, if two burials have both low quantity of glass beads and porcelain in common, the tie is given a value of 2. Our networks are non-directed, which means ties have no orientation forming the relationship between actors to indicate a mutual relationship.

<!-- our hypothesis --> 
Our pericolonial impact hypothesis is that rare, high quality grave goods accumulated in only a few individuals' burials after the European presence in the 17th century, because of increased social inequality at Kiwulan due to differential access to these goods. We can identify these high value trade goods had been from historical documents written by the Europeans and the Han Chinese [@Borao2009; @LiandWu2006]. A related effect of the European presence in northeastern Taiwan may have been manipulation of the European colonial image by ambitious local Indigenous individuals for building their personal status and power. If social inequality gradually increased in Kiwulan as we hypothesize, then we expect to observe a network with higher degree (popularity), less transitivity (cohesion), and strong inequality based on age difference (achieved status). To test this prediction, we use ERGM in a Bayesian framework to examine the formation of network ties and the underlying mechanisms that shape relationships between people at Kiwulan. By comparing networks from the pre-European period and the post-European period we can examine the effects of foreign contact on community relationships at Kiwulan.

## Model specification in a Bayesian framework
```{r ergm-terms-table}
ergm_terms <- readxl::read_excel(here::here("analysis", "data", "raw_data", "ergm_term.xlsx")) 
kableExtra::kable(ergm_terms,
             caption = "The parameters of exponential random graph models for a undirected network to corresponding with archaeological evidence.")
```

<!-- Network construction and specification -->
We used Bayesian inference on exponential random graph models (ERGMs) to quantify the relations among burials and test our hypothesis of social change using the R programming language <!-- cite R --> with the bergm package [@Caimo2014], which is built upon the statnet package for model specification and simulation procedures [@Hunter2008; @Morris2008; @Handcock2008]. Table \@ref(tab:ergm-terms-table) lists the configurations we used for the model specification for burials with the corresponding archaeological evidence. Every parameter in an ERGM has an associated algorithm for computing its value for a network [@Morris2008]. Based on our hypothesis, we model a network with increased social inequality that is represented by low reciprocity, low transitivity, high popularity, high age homophily, and less proximity structure effects. We modeled four node attributes for homophily effects, including age, gender, ritual activity, and the degree of wealth, to explore the correlation between the tendency of connected nodes to form relationships and the network ties overall. We also take into account the physical distance between burials as an indicator of a kinship-based relation since the deceased from the same family tend to be buried nearby [@LiandWu2006]. Our model could reveal the emergence of social inequality via the presence of a few individuals as network centers, having more relations with others. This would be indicated by high popularity or degree values in the network statistics. According to our hypothesis, the burial evidence from after the European arrival will have a better fit with our model compared with the burial evidence from before European contact.

```{r}
burn_in <- 200
...
```

<!--Model fitting and prior setting--> 
After we set our model parameters, we simulated networks in a Bayesian framework using a Markov chain Monte Carlo (MCMC) algorithm. MCMC algorithms allow estimation of posterior distributions through direct random sampling the posterior without assuming the prior comes from any specific distribution [@Hamra2013]. We can obtain a posterior distribution by constructing a Markov chain that describes a sequence of moves from current state to the next state following probabilistic rules based on an algorithm. This enables a random or stochastic simulation in a long run where each move does not depend on the previous move. More chains ensures a more desirable posterior distribution that is close to the target distribution under study, or convergence. In Bayesian ERGMs, MCMC first selects a set of edges (or a set of empty pair of actors) with equal probability, and then switches to a pair of actors at random within the chosen set [@Caimo2011]. In our case, we set the number of chains to six. For each chain, the number of burn-in iterations was 200 `r burn_in` and the number of iterations after the burn-in was 2000. We set the number of iterations used to simulate a network y′ at each iteration to 10,000. We use the normal distribution for the priors, as is typical in network analysis studies that assume networks to have low density and high transitivity, as are commonly found in the real world [@Caimo2017].

<!-- add text about anthropological assumptions of our prior values -->

@@

<!-- Goodness-of-fit assessment-->
The output from our two were models are assessed by goodness-of-fit (GOF) diagnostics in the Bayesian framework, where the observed network is compared with the set of networks simulated from the estimated posterior distributions of the parameters of the model [@Caimo2011; @Caimo2017]. Our Bayesian GOF summarized three distributions, including degree, minimum geodesic distance, and edgewise shared partner distributions. This provides a statistical approach to check how well the estimated posterior parameter distribution based on our hypotheses can reproduce networks similar to the general structural features of the observed network. We then compared the distribution of our observed networks, pre-European and post-European networks, with the distribution of our hypothesized model with higher popularity. We expect to see the network after the European presence has a better fit with the hypothesis, indicative of a structure with higher social differentiation.
